---
title: "4th down research"
author: "Ben Baldwin"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an update of [the analysis that used to be stored here](https://github.com/guga31bb/fourth_calculator/blob/main/R/season_numbers.R). The idea is to show different ways of using the package to analyze 4th down decisions in hopes of inspiring others to try other things.

The code isn't being displayed on this webpage since it is long but [can be accessed here](https://github.com/guga31bb/nfl4th/blob/master/vignettes/articles/4th-down-research.Rmd).

```{r setup, echo = FALSE, message = FALSE}
library(nfl4th)
library(tidyverse)
library(ggplot2)
library(ggtext)
library(DescTools)
library(ggthemes)
library(gt)
library(magick)
library(cowplot)
library(ggpmisc)
```

```{r load}
load_season <- function(season) {
  nflfastR::load_pbp(season) %>%
  nfl4th::add_4th_probs() %>%
  dplyr::mutate(go = ifelse((rush == 1 | pass == 1) & !play_type_nfl %in% c("PUNT", "FIELD_GOAL"), 100, 0)) %>%
  dplyr::filter(down == 4) %>%
  return()
}

pbp_2020 <- load_season(2020)
```


## Worst kick decisions of 2020

```{r worst, echo = FALSE}
pbp_2020 %>%
  filter(
    go == 0,
    # they tried to go for it
    !(posteam == "ARI" & week == 3 & play_id == 2364)
  ) %>%
  mutate(
    defteam = if_else(posteam == home_team, away_team, home_team)
  ) %>%
  arrange(-go_boost) %>%
  mutate(rank = 1 : n()) %>%
  head(10) %>%
  select(rank, posteam, defteam, week, qtr, ydstogo, diff = score_differential, go_boost, desc) %>%
  gt() %>%
  cols_label(
    rank = "",
    posteam = "Team",
    defteam = "Opp",
    week = "Week",
    qtr = "Qtr",
    ydstogo = "YTG",
    diff = "Diff",
    desc = "Play",
    go_boost = "WP loss"
  ) %>%
  tab_style(
    style = cell_text(color = "black", weight = "bold"),
    locations = list(
      cells_row_groups(),
      cells_column_labels(everything())
    )
  ) %>% 
  text_transform(
    locations = cells_body(vars(posteam)),
    fn = function(x) web_image(url = paste0('https://a.espncdn.com/i/teamlogos/nfl/500/',x,'.png'))
  ) %>% 
  text_transform(
    locations = cells_body(vars(defteam)),
    fn = function(x) web_image(url = paste0('https://a.espncdn.com/i/teamlogos/nfl/500/',x,'.png'))
  ) %>% 
  cols_width(
    everything() ~ px(400),
    ) %>% 
  cols_width(
    vars(rank) ~ px(30),
    vars(posteam) ~ px(50),
    vars(defteam) ~ px(50),
    vars(week) ~ px(50),
    vars(diff) ~ px(50),
    vars(qtr) ~ px(50),
    vars(ydstogo) ~ px(50),
    vars(go_boost) ~ px(70)
  ) %>% 
  tab_options(
    row_group.border.top.width = px(3),
    row_group.border.top.color = "black",
    row_group.border.bottom.color = "black",
    table_body.hlines.color = "white",
    table.border.top.color = "black",
    table.border.top.width = px(1),
    table.border.bottom.color = "white",
    table.border.bottom.width = px(1),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(2),
    row.striping.background_color = '#FFFFFF',
    row.striping.include_table_body = TRUE,
    table.background.color = '#F2F2F2',
    data_row.padding = gt::px(2),
    table.font.size = gt::px(16L)
  ) %>%
  fmt_number(
    columns = vars(go_boost), decimals = 1
  ) %>%
  cols_align(
    columns = 1:8,
    align = "center"
  ) %>% 
  tab_header(
    title = md(glue::glue("Worst kick decisions of 2020"))
  )
```

## Leaguewide behavior by strength of recommendation
```{r league, echo = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 800}
theme_ben <- theme_fivethirtyeight() +
  theme(
    legend.position = "none",
    plot.title = element_markdown(size = 22, hjust = 0.5),
    plot.subtitle = element_markdown(size = 12, hjust = 0.5),
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(size=12, face="bold")
  )

plot <- pbp_2020 %>%
  mutate(go_boost = RoundTo(go_boost, 0.5)) %>%
  group_by(go_boost) %>%
  summarize(go = mean(go)) %>%
  ungroup() %>%
  filter(between(go_boost, -10, 10)) %>%
  mutate(
    should_go = case_when(go_boost > .5 ~ 1,
                          go_boost < -.5 ~ 0,
                          TRUE ~ 2)
  )

plot %>%
  ggplot(aes(go_boost, go, color = as.factor(should_go))) + 
  geom_point(size = 5, color = "black", alpha = .5) +
  geom_vline(xintercept = 0) +
  geom_smooth(method = "lm", show.legend = F, se = F, size = 3)+
  theme_ben +
  labs(x = "Gain in win probability by going for it",
       y = "Go-for-it percentage",
       caption = paste0("Figure: @benbbaldwin"),
       subtitle = "By strength of @ben_bot_baldwin recommendation, 2020",
       title = glue::glue("NFL Go-for-it Rate on <span style='color:red'>4th down</span>")
       ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0,2)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 20), limits = c(-10, 10), expand = c(0,0)) +
  annotate("text",x=-4, y= 90, label = "Should\nkick", color="red", size = 5) +
  annotate("text",x=3, y= 90, label = "Should\ngo for it", color="red", size = 5) +
  annotate("label",x=-6, y= 15, label = "Teams almost always kick\nwhen they should...", size = 5) +
  annotate("label",x=6, y= 25, label = "...but frequently\n kick when they\nshould go for it", size = 5)
```

## League behavior based on win probability

```{r wp, echo = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 800}

min <- 2
max <- 4

my_title <- glue::glue("NFL Go-for-it Rate on <span style='color:red'>4th down</span>")

pbp_2020 %>%
  filter(go_boost > min & go_boost < max) %>%
  mutate(vegas_wp = 100 * vegas_wp) %>%
  ggplot(aes(vegas_wp, go)) + 
  # geom_point(size = 5, color = "black", alpha = .5) +
  geom_smooth(show.legend = F, se = F, size = 3, color = "black")+
  theme_ben +
  labs(x = "Win probability prior to play",
       y = "Go-for-it percentage",
       caption = paste0("Figure: @benbbaldwin | 2020 season"),
       subtitle = glue::glue("@ben_bot_baldwin gain in win prob by going {min}-{max} percentage points"),
       title = my_title) +
  theme(
    legend.position = "none",
    plot.title = element_markdown(size = 22, hjust = 0.5),
    plot.subtitle = element_markdown(size = 12, hjust = 0.5)
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(0, 100)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 20)) 
```

## Decisions by team
``` {r divs, echo = FALSE, message = FALSE}
divs <- teamcolors::teamcolors %>%
  filter(league == "nfl") %>%
  select(name, division) %>%
  mutate(name = case_when(
    name == "Washington Redskins" ~ "Washington Football Team",
    name == "Oakland Raiders" ~ "Las Vegas Raiders",
    TRUE ~ name
  ))

nfl_logos_df <- nflfastR::teams_colors_logos %>%
  filter(!(team_abbr %in% c("LAR", "SD", "OAK"))) %>%
  left_join(divs, by = c("team_name" = "name"))

follow_bot <- function(conf) {
  
  logos_conf <- image_read(nfl_logos_df %>% filter(substr(division, 1, 3) == conf) %>% pull(team_logo_espn))
  my_title <- glue::glue("<span style='color:red'>{conf}</span>: when did teams follow the bot? 2020")
  
  pbp_2020 %>% 
    left_join(nfl_logos_df, by = c("posteam" = "team_abbr")) %>%
    filter(substr(division, 1, 3) == conf) %>%
    filter(between(go_boost, -5, 15)) %>%
    mutate(go = as_factor(go)) %>%
    select(game_id, posteam, team_logo_espn, go, go_boost, yardline_100, score_differential, ydstogo, desc) %>%
    ggplot(aes(x = posteam, y=go_boost, color=go)) +
    geom_hline(yintercept = 0) +
    geom_jitter(aes(y = go_boost, fill = go), 
                size = 3, width = 0.2, show.legend=FALSE, alpha=.5) +
    scale_y_continuous(name = "4th down bot gain in going for it", breaks = scales::pretty_breaks(n = 5), expand = c(0,.03)) +
    theme_ben +
    labs(title = my_title,
         x = "Went for it",
         subtitle = "<span style='color:#F8766D'>Kicked</span> versus <span style='color:#00BFC4'>went for it</span>",
         caption = "Predicted go based on past league behavior given\nyardage, time, score, field position, week, and season"
    ) +
    theme(
      panel.grid.major = element_blank(),
      plot.subtitle = element_markdown(size = 16, hjust = 0.5),
      axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank()
    ) +
    geom_hline(yintercept = -6, color = "gray", alpha = .1) +
    # doing this in a smart way is annoying so just brute forcing it
    draw_image(logos_conf[1],  x = .5, y = -5.75, scale = 2) +
    draw_image(logos_conf[2],  x = 1.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[3],  x = 2.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[4],  x = 3.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[5],  x = 4.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[6],  x = 5.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[7],  x = 6.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[8],  x = 7.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[9],  x = 8.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[10],  x = 9.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[11],  x = 10.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[12],  x = 11.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[13],  x = 12.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[14],  x = 13.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[15],  x = 14.5, y = -5.75, scale = 2) +
    draw_image(logos_conf[16],  x = 15.5, y = -5.75, scale = 2)
  
}
```

```{r afc, echo = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 800}
follow_bot("AFC")
```

```{r nfc, echo = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 800}
follow_bot("NFC")
```

```{r, echo = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 800}
current <- pbp_2020 %>%
  filter(go_boost > 1.5) %>%
  filter(vegas_wp > .2) %>%
  group_by(posteam) %>%
  summarize(go = mean(go), n = n()) %>%
  ungroup() %>%
  left_join(nflfastR::teams_colors_logos, by=c('posteam' = 'team_abbr')) %>%
  arrange(-go) %>%
  mutate(rank = 1:n()) %>%
  arrange(posteam)

ids <- nflfastR::teams_colors_logos %>%
  filter(!team_abbr %in% c('LAR', 'OAK', 'SD', 'STL'))
images <- magick::image_read(ids%>% pull(team_logo_espn)) 

logos <- tibble(
  x = current$rank + .25, 
  y = current$go + .02,
  width = .035,
  grob = 
    map(1:32, function(x) {
      grid::rasterGrob(images[x])
    })
)

my_title <- glue::glue("Which teams <span style='color:red'>go for it</span> when they <span style='color:red'>should?</span> {max(pbp_2020$season)}")
ggplot(data = current, aes(x = reorder(posteam, -go), y = go)) +
  geom_col(data = current, aes(fill = ifelse(posteam=="SEA", team_color2, team_color)), 
           width = 0.5, alpha = .6, show.legend = FALSE
  ) +
  geom_grob(data = logos, 
            aes(x, y, label = grob, vp.width = width),
            hjust = 0.7) +
  scale_fill_identity(aesthetics = c("fill", "colour")) +
  theme_ben +
  theme(
    panel.grid.major.x = element_blank(),
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank()
    ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  labs(
    x = "",
    y = "Go rate",
    title= my_title,
    subtitle = "When @ben_bot_baldwin recommends going for it (gain in win prob. at least 1.5 percentage points)",
    caption = glue::glue("Sample size in parentheses\nExcl. final 30 seconds of game. Win prob >20%")
  ) +
  geom_text(data = current, aes(x = rank, y = -.015, size=.03, label = glue::glue("({n})")), show.legend = FALSE, nudge_x = 0, color="black")
```

```{r, echo = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 800}
current <- pbp_2020 %>%
  group_by(posteam) %>%
  mutate(
    games = n_distinct(game_id),
  ) %>%
  ungroup() %>%
  filter(go_boost > 0, go == 0) %>%
  group_by(posteam) %>%
  summarize(
    go = sum(go_boost), 
    n = n(),
    games = dplyr::first(games),
    go = go/games
    ) %>%
  ungroup() %>%
  left_join(nflfastR::teams_colors_logos, by=c('posteam' = 'team_abbr')) %>%
  arrange(-go) %>%
  mutate(rank = 1:n()) %>%
  arrange(posteam)

ids <- nflfastR::teams_colors_logos %>%
  filter(!team_abbr %in% c('LAR', 'OAK', 'SD', 'STL')) %>%
  filter(team_abbr %in% current$posteam)
images <- magick::image_read(ids%>% pull(team_logo_espn)) 

logos <- tibble(
  x = current$rank + .25, 
  y = current$go + .02,
  width = .035,
  grob = 
    map(1:length(images), function(x) {
      grid::rasterGrob(images[x])
    })
)

my_title <- glue::glue("WP per game <span style='color:red'>lost by kicking in go situations</span>, {max(pbp_2020$season)}")
ggplot(data = current, aes(x = reorder(posteam, -go), y = go)) +
  geom_col(data = current, aes(fill = ifelse(posteam=="SEA", team_color2, team_color)), 
           width = 0.5, alpha = .6, show.legend = FALSE
  ) +
  geom_grob(data = logos, 
            aes(x, y, label = grob, vp.width = width),
            hjust = 0.7) +
  scale_fill_identity(aesthetics = c("fill", "colour")) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_markdown(size=20,face = 2,hjust=.5),
    plot.subtitle = element_text(size=8, hjust=.5),
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank()
  ) +
  # scale_y_continuous(expand=c(0,0), limits=c(0, max(current$go + 5))) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    x = "",
    y = "Win probability lost per game",
    title= my_title,
    caption = glue::glue("@benbbaldwin | Excl. final 30 seconds of game")
  )
```
